|

### 2022-07-24 manuscript.Rmd
add chunk name
expand data viz section

bibliography: ["bib/references.bib", "bib/packages.bib"]
output:
  rticles::jds_article: default
  bookdown::pdf_book:
    base_format: rticles::jds_article # for using bookdown features like \@ref()
---

@@

```{r pkg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, purl=FALSE}
# Load required packages
## Any packages loaded in this chunk will automatically be added to `bib/packages.bib` for citation use
library(ggplot2)
library(scales)
library(BrailleR)
library(sonify)
library(tactileR)
library(tidyverse)
library(shiny)
library(knitr)
library(gt)
```

@@

Data visualization is the most common representation of data. 
Colors are an important part of accessibility of data visualizations. 
In our courses, while teaching colors in data visualizations, we introduce color-blindness through simulators. 
Those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness. 
Students get to experience this for red-, green-, and blue-blindness. 
For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe]. 
In addition we rely on `ggplot2::scale_fill_viridis_d()` for generating `ggplot` visualizations [@ggplot22016] that can be perceived by those with common forms of color blindness.

@@

```{r include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
knitr::write_bib(file = "bib/packages.bib")
```

>>>>

bibliography: ["bib/references.bib", "bib/packages.bib"]
output:
  rticles::jds_article: 
    includes:
      in_header: header.tex #header-hidden.tex can be used when creating an anonymized version for review
  bookdown::pdf_book:
    base_format: rticles::jds_article # for using bookdown features like \@ref()
---

@@

```{r pkg, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, purl=FALSE}
# Load required packages
## Any packages loaded in this chunk will automatically be added to `bib/packages.bib` for citation use
library(ggplot2)
library(scales)
library(BrailleR)
library(sonify)
library(tactileR)
library(tidyverse)
library(shiny)
library(knitr)
library(gt)
library(colorblindr)
library(palmerpenguins)
```

@@

Data visualization is the most common representation of data and it is covered in many introductory data science curricula and textbooks <!--Mine will add citations-->.
Despite their use by many, data visualizations are not accessible to all. 
Colors are an important part of data visualizations and an aspect that can limit accessibility for color blind people. 
When teaching data visualizations we include three basic rules for incorporating colors into data visualization. Students should be able to

1. simulate color blindness (if they are sighted and not color blind);
2. pick color blind friendly colors.
3. not only rely on color for differentiating data;

We show an example of incorporating these three basic rules by using the Palmer Penguins data <!--Mine will cite this-->. 
We visualize the relationship between flipper length and bill length of penguins and color the data points differently for each species in Figure \ref{fig:penguins-basic}.

```{r penguins-basic, echo=FALSE, fig.cap="Visualization showing the relationship between flipper length (mm) and bill length (mm) of penguins for three different species of penguins", fig.alt="", warning=FALSE, fig.height=3}
library(ggplot2) #It seems like the pkg chunk is not loading packages. I did not want to make changes there since I do not know what your intention was.
library(palmerpenguins)
fig <- ggplot(penguins, aes(flipper_length_mm, bill_length_mm, color = species)) +
         geom_point() +
  labs(x = "Flipper Length(mm)",
       y = "Bill Length (mm)",
       color = "Species") +
  theme_bw()

fig

```

The `colorblindr` package provides a set of functions that can be useful in teaching and practising color blindness accessibility. 
For instance, `cvd_grid()` function provides color-deficiency simulations of a given plot. 
In the code chunk below the plot is named as `fig`. 
The code creates a grid of four plots where each plot shows a different color vision deficiency.
This grid allows those without any color vision deficiencies experience what data visualizations would look like to a sighted person that does have color blindness.

```{r penguins-cvd, echo=FALSE, fig.cap="", fig.alt="", warning=FALSE}
colorblindr::cvd_grid(fig)
```

For their data visualizations, instead of relying on the default color selection of R packages, students are introduced to the Okabe-Ito color palette which is known to be accessible to people with color vision deficiencies [@okabe]. 
One can find specific colors in the pallette by the R code `palette.colors(palette = "Okabe-Ito")`.
In addition, students can be introduced to `colorblindr::scale_color_OkabeIto()` and `colorblindr::scale_fill_OkabeIto()` to automatically switch to using this color pallette by adding a layer to a ggplot presentation. 
For instance, the plot in Figure \ref{fig:penguins-basic} can be switched to Okabe-Ito color pallette as shown in Figure \ref{fig:penguins-okabe-ito} by using the code below.

```{r penguins-okabe-ito, fig.cap="", fig.alt="", warning=FALSE, fig.height=3}
fig + colorblindr::scale_color_OkabeIto()
```

Last but not least, students should learn not to rely only on color to differentiate data points, in this case Species. 
For instance, they can be introduced to shapes and have data points that are circle, triangle, square for each species. 
This can also be extended to line graphs as having continuous or dashed lines.
Facetting can also be utilized. 
For instance, each of the species could have its own scatterplot side-by-side.

In this manuscript and in our introductory courses, we focus on static data visualizations. 
Those who teach interactive data visualizations may also consider covering cognitive and motor accessibility. 
Our focus here is on visual accessibility but it is worth noting that all subdomains of accessibility are relevant to data science learners and can be incorporated into the curriculum as seen fit.

@@

```{r pkg_bib, include=FALSE, echo=FALSE, warning=FALSE, message=FALSE, cache=TRUE, dependson="pkg"}
knitr::write_bib(file = "bib/packages.bib")
```
